Finally tagless, partially evaluated

### 2007-04-06 tagless/tagless-final.tex
Related work done

\hyphenation{meta-language meta-languages Meta-OCaml meta-cir-cu-lar type-case}
\documentclass[preprint]{sigplanconf}
\usepackage{amsmath,amssymb}
\usepackage{mathptmx}
%\usepackage{txfonts} \renewcommand{\ttdefault}{cmtt}
\usepackage[T1]{fontenc}
\usepackage[override]{cmtt}
\usepackage{comment}
\usepackage{hyphenat}
\usepackage{url}
\usepackage{prooftree1} \proofrulebaseline=1.5ex \proofdotseparation=.75ex
\usepackage{mdwlist}
\usepackage{mdwtab}
\usepackage{dblfloatfix}

@@

\title{Finally tagless, partially evaluated}
\subtitle{Tagless Staged Interpreters for Simpler Typed Languages}
%Staged typed type-preserving tagless ``interpreters'', \emph{finally}

@@

\subsection{Contexts clarify polymorphism}

@@

\citet{WalidICFP02}, when referring to the work of \citet{taha-tag},
claims that this shows that self-interpretation is possible in a
\emph{simply typed $\lambda$-calculus}, but this is somewhat misleading
as their type system does not differentiate between the various ground
types.  In other words, various constants (like \textsf{nil} and 
\textsf{true}) both inhabit the same type $D$.  This is not very satisfying.
Furthermore, as is pointed out in the introduction to \citet{WalidICFP02},
\citet{taha-tag} does not \emph{statically} guarantee that all tags will
be removed - tag elimination must be performed at runtime.

\Citet{WalidICFP02}, \citet{taha-tag}, \citet{xi-guarded}, and
\citet{peyton-jones-simple} seem to argue as follows that a self\hyp
interpreter of a typed language cannot be Jones\hyp optimal.
\begin{enumerate*}
\item One needs to encode a typed language in a typed language based on
a sum type (at some level of the hierarchy),
\item It is not possible to write a \emph{direct} interpreter 
for such an encoding of a typed language
in a typed language, without either an
advanced type system or the overhead of tagging and untagging.
\item Thus an indirect interpreter is necessary, which needs a universal
  type (and hence tagging),
\item Thus any self-interpreter must have tags and hence cannot be 
  Jones-optimal.
\end{enumerate*}
While the logic is sound, we showed that the premise in the very first step
was not valid.

\citet{Danvy-tagging-encoding} discuss Jones optimality at length and
demonstrate the usefulness of HOAS for typed self-interpretation.
However, no claim is made that the source program is itself typed --
just that the interpreter is.  Our reading is that the source language
is not typed. Therefore, object term encoding has tags and the
interpreter (given in the code accompanying
\citet{Danvy-tagging-encoding}) extensively uses partial
pattern-matching. The shift to HOAS, however, lets the partial
evaluator to remove all the tags. In contrast, our object encoding and
evaluators do not have tags to start with and have no possibility of
pattern-match errors.

We used a method from \citet{asai-binding-time} to reduce the amount
of back-and-forth between reify/reflect: in our PE we track both
static and dynamic values in parallel.  In some sense,
we are observing that this method type-checks in Hindley-Milner once we
deforest the static code representation.  Put another way, we have
manually applied type-level partial-evaluation to our typelevel 
functions (see \S\ref{S:PE-solution}) to obtain simpler types 
acceptable to MetaOCaml.  It is worthwhile to note that 
\citet{sumii-hybrid} also takes advantage of this method
\citep[of][]{asai-binding-time} to get interesting results combining online
and offline partial evaluation.  Like \citet{SperberThiemann:TwoForOne},
we strive for modularity by reusing earlier stages.  It would be interesting
to see if we could derive a \emph{cogen} \citep{Thiemann:cogeninsixlines}
in the same manner.

The approach to HOAS via catamorphism and anamorphism of
\citet{Washburn-Weirich-boxes} seems related to what we are doing.
We have not investigated the relationship to any serious depth, but we
feel that one could fruitfully combine their ideas with ours.

A lot of effort has gone into finding ways to give precise types to untyped
terms \citep{baars-typing,Guillemette-Monier-PLPV,haskell-list}).  These 
take advantage of the host language's types to varying extents.  While
very successful for proper terms, results for
ill-typed terms can produce rather obscure error messages.  We use the
host's typing facilities for this, and thus inherit the strengths
and weaknesses of the host language.  While that may not be optimal, it is
certainly much more modular.

When dealing with datatypes which are equivalent but have different
representations, it is customary to use injection-projection pairs
(used in the context of interpreters by \citet{Ramsey-ML-module-mania}
and \citet{Benton-embedded-interpreters}, partial evaluation by
\citet{Danvy-TDPE}, and type-level functions by
\citet{oliveira-typecase}).  Na\"{\i}vely, this would appear to apply to
our implementation of the type-level function |'sv -> 'dv|.  However, in most
of these cases, it is important that this establishes a bijection (between
|'sv| and |'dv| in our case).  Our approach is one-sided: we requite
an operational definition of |'sv -> 'dv| (which is easy), and an
implementation of |'dv| which ``perfectly reflects'' |'sv| but without 
requiring an implementation.

We are not the first to implement a typed interpreter for a typed
language -- \citet{laod93} use a typed version of the SK language
(and typeclasses) to implement a meta-circular interpreter (rather than
a self-interpreter).  Interestingly, their meta-circular interpreter
appears to be tagfree!  However, they could not have implemented a
compiler or a partial-evaluator in the same way since they rely quite
heavily on having injection-projection pairs.

\citet{fiore:nbe-ppdp2002,balat:tdpe-popl2004} also deal with a 
tagfree partial-evaluator using quite a lot of interesting 
technology (amongst them powerful continuation-based control operators).
While they use a reify-reflect injection-projection pair, this appears
to be done ``lightly'', in the sense that the structure of compiled code
is never examined.  The exact relation between their work and ours is 
still a little mysterious to us, but we will continue to investigate it.

We have been unable to find work that establishes that
the \emph{typed} $\lambda$-calculus possesses a final coalgebra structure.
For the untyped $\lambda$-calculus, 
\citet{HonsellLenisa,honsell99coinductive} establish several interesting
results along this line.  Of particular interest is their use of
contexts with a hole \citep[p.13]{honsell99coinductive} for the definition
of \emph{observational equivalence}
(see \S\ref{selfinterp}).  The reader is directed to bibliography of
\citet{honsell99coinductive} for references to the foundational work
in this important area.  Particularly intriguing is the link to the
coinductive aspects of B\"{o}hm trees, as pointed out by
\citet{berarducci-models} and \citet[Example 4.3.4]{jacobs-coalgebra}.

One way to interpret some of our results is that we eschew sums (at either the
term or type, or even kind level) for our terms, but instead use a record type 
(via Functors in OCaml and typeclasses in Haskell).
For the self-interpreter, we then proceed to use a Church encoding for
recursive data types \citep{bohm-automatic}.

>>>>

\listfiles
\hyphenation{meta-language meta-languages Meta-OCaml meta-cir-cu-lar type-case}
\documentclass[preprint]{sigplanconf}
\usepackage{amsmath,amssymb}
\usepackage{mathptmx}
%\usepackage{txfonts} \renewcommand{\ttdefault}{cmtt}
\usepackage[T1]{fontenc}
\usepackage[override]{cmtt}
\usepackage{comment}
\usepackage{hyphenat}
\usepackage{url}
\usepackage{prooftree1} \proofrulebaseline=1.5ex \proofdotseparation=.75ex
\usepackage{mdwlist}
\usepackage{mdwtab}
\usepackage{dblfloatfix}

@@

\title{Finally tagless, partially evaluated}
\subtitle{Tagless staged interpreters for simpler typed languages}
%Staged typed type-preserving tagless ``interpreters'', \emph{finally}

@@

\subsection{Contexts clarify polymorphism}
\label{S:clarify}

@@

Even a simply typed $\lambda$-calculus obviously supports self\hyp
interpretation, provided we use universal types \citep{taha-tag}.  The
ensuing tagging overhead motivated \citet{taha-tag} to propose tag
elimination, which however does not statically guarantee that all tags
will be removed \citep{WalidICFP02}.

\Citet{WalidICFP02}, \citet{taha-tag}, \citet{xi-guarded}, and
\citet{peyton-jones-simple} seem to argue as follows that a self\hyp
interpreter of a typed language cannot be tagless or Jones\hyp optimal.
\begin{enumerate*}
\item One needs to encode a typed language in a typed language based on
a sum type (at some level of the hierarchy).
\item It is not possible to write a \emph{direct} interpreter 
for such an encoding of a typed language
in a typed language, without either an
advanced type system or the overhead of tagging and untagging.
\item Thus an indirect interpreter is necessary, which needs a universal
  type (and hence tagging).
\item Thus any self-interpreter must have tags and cannot be 
  Jones-optimal.
\end{enumerate*}
While the logic is sound, we showed that the premise in the very first step
was not valid.

\citet{Danvy-tagging-encoding} discuss Jones optimality at length and
apply HOAS to typed self\hyp interpretation.  However, their source
language is untyped.  Therefore, their object\hyp term encoding has
tags, and their interpreter can raise run-time errors.
Nevertheless, HOAS lets the partial
evaluator remove all the tags. In contrast, our object encoding and
interpreters do not have tags to start with and obviously cannot
raise run-time errors.

Our final approach seems related to \citearound{'s approach to HOAS via
catamorphism and anamorphism}\citet{Washburn-Weirich-boxes}.

A lot of effort has gone into ``typing dynamic typing'': to statically
type-check dynamically\hyp typed values
\citep{baars-typing,WalidICFP02,Guillemette-Monier-PLPV,haskell-list},
using the host language's type system to varying extents.
Our object terms are statically typed, so we would
need one of these techniques to interpret dynamically\hyp typed
terms such as those read from a file.

Our partial evaluator establishes a bijection |repr_pe| between static
and dynamic types (the valid values of |'sv| and |'dv|), and between
static and dynamic terms.  It is customary to implement such a bijection
using an injection\hyp projection pair, as done for interpreters by
\citet{Ramsey-ML-module-mania} and \citet{Benton-embedded-interpreters},
partial evaluation by \citet{Danvy-TDPE}, and type-level functions by
\citet{oliveira-typecase}.  As explained in~\S\ref{S:PE-solution}, we
avoid injection and projection at the type level by adding an argument
to |repr|.

At the term level, we also avoid converting between static and dynamic
terms by building them in parallel, using \citearound{'s
method}\citet{asai-binding-time}.
This method type-checks in Hindley-Milner once we
deforest the object term representation.  Put another way, we
manual apply type-level partial evaluation to our type
functions (see \S\ref{S:PE-solution}) to obtain simpler types 
acceptable to MetaOCaml.
\Citet{sumii-hybrid} also use Asai's method,
to combine online
and offline partial evaluation.  Like \citet{SperberThiemann:TwoForOne},
we strive for modularity by reusing interpreters for individual stages.  It would be interesting
to try to derive a \emph{cogen} \citep{Thiemann:cogeninsixlines}
in the same manner.

We are not the first to implement a typed interpreter for a typed
language.  \Citet{laod93} use type classes to implement a metacircular
interpreter (rather than a self\hyp interpreter) of a
typed version of the SK language, which is quite different from our
object language.  Their interpreter
appears to be tagless, but they could not have implemented a
compiler or partial evaluator in the same way, since they rely
heavily on injection\hyp projection pairs.

\Citet{fiore:nbe-ppdp2002} and \citet{balat:tdpe-popl2004} also build
a tagless partial evaluator, using delimited control operators.  It is
type-directed, so the user must represent, as a term, the type of every
term to be partially evaluated.  We shift this work to the type checker
of the metalanguage.  By avoiding term-level type representations, our
approach makes it easier to perform algebraic simplifications (as
in~\S\ref{S:PE-solution}).

We could not find work that establishes that
the \emph{typed} $\lambda$-calculus possesses a final coalgebra structure.
\Citet{HonsellLenisa,honsell99coinductive}
investigate the untyped $\lambda$-calculus,
along this line.  In particular, they use
contexts with a hole \citep[p.\,13]{honsell99coinductive} to define
\emph{observational equivalence}
(see \S\ref{selfinterp}).
\citearound{'s bibliography}\Citet{honsell99coinductive} refers to the
foundational work in this important area.  Particularly intriguing is the link to the
coinductive aspects of B\"{o}hm trees, as pointed out by
\citet{berarducci-models} and \citet[Example 4.3.4]{jacobs-coalgebra}.

One way to understand our main idea is to eschew sum types and sum kinds
for their dual, record types and record kinds.
For the self\hyp interpreter, we then proceed to use a Church encoding for
recursive data types \citep{bohm-automatic}.

As \S\ref{S:clarify} observes, higher-rank and higher-kind
polymorphism lets us type-check and compile object terms separately from
interpreters.  This observation is consistent with the role of
polymorphism in the separate compilation of modules
\citep{shao-typed,shao-transparent}.
